### 2020-10-22 papers/thesis/full/ch2-spherical-harmonics/spherical-harmonics.tex
New thesis structure - commiting important files

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or epsÂ§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex

\input{../includes/bsmacros}
\input{../includes/bsthesisnewcommands}
\input{../includes/somacros}

\title{Spherical harmonics as orthogonal polynomials in three variables}
\author{Ben Snowball}
%\date{}							% Activate to display a given date or no date

\begin{document}

\maketitle

\section{Spherical harmonics as orthogonal polynomials in three variables}

\subsection{Introduction}

@@

\subsection{Defining the spherical harmonics in three variables}

@@

@@

\begin{proof}
	The recurrence relationship for the Jacobi polynomials satisfies \cite[18.9.1, 18.9.2]{DLMF} 
\bseqn{
	z P^{(m,m)}_{l-m}(z) = \tilde \gamma_{l,m,1} P^{(m,m)}_{l-m-1}(z) + \tilde \gamma_{l,m,2} P^{(m,m)}_{l-m+1}(z),
}
for $k \ge 0, \, m \in \Z$. Hence, the expression for multiplication of the spherical harmonic $\Ylmfull$ by $z$ is
\bseqn{
	z\,\Ylm(x,y,z) &= \clm \eimphi \rho(z)^{|m|} z P^{(|m|,|m|)}_{l-|m|}(z)  \\
		     &= \clm \eimphi \rho(z)^{|m|} \big[ \tilde \gamma_{l,m,1} P^{(|m|,|m|)}_{l-|m|-1}(z) + \tilde \gamma_{l,m,2} P^{(|m|,|m|)}_{l-|m|+1}(z) \big] \\
		     &= \gamma_{l,m,1} Y^{m}_{l-1}(x,y,z) + \gamma_{l,m,2} Y^{m}_{l+1}(x,y,z).
}

For multiplication by $x$ and $y$, we require some further relations for the complex exponential and the Jacobi polynomials. First, recall that
\bseqn{
	\cosphi \, \eimphi &= \frac{1}{2} (e^{i\varphi} + e^{-i\varphi}) \eimphi =  \frac{1}{2} (e^{i(m+1)\varphi} + e^{i(m-1)\varphi}) \\
	\sinphi \, \eimphi &= \frac{1}{2i} (e^{i\varphi} - e^{-i\varphi}) \eimphi =  \frac{-i}{2} (e^{i(m+1)\varphi} - e^{i(m-1)\varphi})
}
Finally, we can express a Jacobi polynomial $P^{(a,b)}_{l-m}(z)$ in terms of an ultraspherical polynomial $C^{(\lambda)}_{l-m}(z)$ \cite[18.7.2]{DLMF}:
\bseqn{
	{l! \: (2m)! \over (l+m)! \: m!} \: C^{(m+1/2)}_{l-m}(z) = P^{(m,m)}_{l-m}(z),
}
where the ultraspherical polynomials satisfy the relations \cite[18.9.7, 18.9.8]{DLMF}:
\bseqn{
	C^{(\lambda)}_n(z) &= {\lambda \over n + \lambda} \: \big[ C^{(\lambda+1)}_n(z) - C^{(\lambda+1)}_{n-2}(z) \big]  \\
	(1-z^2) \, C^{(\lambda)}_n(z) &= {1 \over 4(\lambda-1)(n+\lambda)} \: \big[ (n+2\lambda-2)(n+2\lambda-1) \: C^{(\lambda-1)}_n(z) \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad - (n+1)(n+2) \: C^{(\lambda-1)}_{n+2}(z) \big] ,
}
for $m \in \No$ where $\lambda > -\half$ is some parameter. Thus, combining these, we can write three-term recurrences for the Jacobi polynomials as:
\bseqn{
	P^{(m,m)}_{l-m}(z) &= \tilde \alpha_{l,m,4} P^{(m+1,m+1)}_{l-m}(z) + \tilde \alpha_{l,m,2} P^{(m+1,m+1)}_{l-m-2}(z) \\
	(1-z^2) \, P^{(m,m)}_{l-m}(z) &= \tilde \alpha_{l,m,3} P^{(m-1,m-1)}_{l-m+2}(z) + \tilde \alpha_{l,m,1} P^{(m-1,m-1)}_{l-m}(z),
}
for $l, m \in \No$ s.t. $0 \le m \le l$. Hence, the expressions for multiplication of the spherical harmonic $\Ylmfull$ by $x$ and $y$ is then:
\bseqn{
	x\,\Ylm(x,y,z) &= \clm \cosphi \: \eimphi \: \sintheta \: \rho(z)^{|m|} P^{(|m|,|m|)}_{l-|m|}(z) \\
	&= \half \clm (e^{i(m+1)\varphi} + e^{i(m-1)\varphi}) \rho(z)^{|m|+1} P^{(|m|,|m|)}_{l-|m|}(z) \\
	&= \half \clm e^{i(m+1)\varphi} \rho(z)^{|m|+1} \: \big[ \tilde \alpha_{l,m,4} P^{(|m|+1,|m|+1)}_{l-|m|}(z) + \tilde \alpha_{l,m,2} P^{(|m|+1,|m|+1)}_{l-|m|-2}(z) \big] \\
	&\quad \quad + \half \clm e^{i(m-1)\varphi} \rho(z)^{|m|-1} \: \big[ \tilde \alpha_{l,m,3} P^{(|m|-1,|m|-1)}_{l-|m|+2}(z) + \tilde \alpha_{l,m,1} P^{(|m|-1,|m|-1)}_{l-|m|}(z) \big] \\
	&= \alpha_{l,m,1} Y^{m-1}_{l-1}(x,y,z) +  \alpha_{l,m,2} Y^{m+1}_{l-1}(x,y,z) \\
	&\quad \quad + \alpha_{l,m,3} Y^{m-1}_{l+1}(x,y,z) + \alpha_{l,m,4} Y^{m+1}_{l+1}(x,y,z), \\
	y\,\Ylm(x,y,z) &= \clm \sinphi \: \eimphi \: \sintheta \: \rho(z)^{|m|} P^{(|m|,|m|)}_{l-|m|}(z) \\
	&= -\half i \: \clm (e^{i(m+1)\varphi} - e^{i(m-1)\varphi}) \: \rho(z)^{|m|+1} P^{(|m|,|m|)}_{l-|m|}(z) \\
	&= - i \: \big[\Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \big] \\
	&\quad \quad \quad + i \, \big[ \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z) \big], \\
	&= i \: \Big\{ \alpha_{l,m,1} Y^{m-1}_{l-1}(x,y,z) - \alpha_{l,m,2} Y^{m+1}_{l-1}(x,y,z) \\
	&\quad \quad \quad + \alpha_{l,m,3} Y^{m-1}_{l+1}(x,y,z) - \alpha_{l,m,4} Y^{m+1}_{l+1}(x,y,z) \Big\}.
}

@@

\subsection{Three-term recurrence relation for $\SHvec$}

Three-term recurrence relations for orthogonal polynomials are well established (e.g. \cite[18.9]{DLMF}, \cite{dunkl2014orthogonal}). In a similar vein, we can obtain a recurrence relation for the spherical harmonics by combining each system in (\ref{eqn:jacobimatsdef}).
\bseqn{
	\renewcommand\arraystretch{1.3}
	\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y,z) & A_0 & & \\
		C_1 & B_1-G_1(x,y,z) & \quad A_1 \quad & \\
		& C_2 & B_2 - G_2(x,y,z)  & \ddots \\
		& & \ddots &\ddots
	\end{pmatrix}
	\SHvec(x,y,z)
	=
	\begin{pmatrix}
		Y_0 \\ 0 \\ 0 \\ 0 \\ \vdots  \\
	\end{pmatrix},
}
where we note $Y_0^0(x,y,z) \equiv Y_0 := P_0^{(0,0)} \equiv 1$, and for each $l = 0,1,2\dots$,
\begin{align}
	A_l &:= 
		\begin{pmatrix}
			A_{x,l} \\
			A_{y,l} \\
			A_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l+3)}, \quad
	C_l := 
		\begin{pmatrix}
			C_{x,l} \\
			C_{y,l} \\
			C_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l-1)} \quad (n \ne 0), \label{eqn:clenshawmats1} \\
	B_l &:= 
		\begin{pmatrix}
			B_{x,l} \\
			B_{y,l} \\
			B_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l+1)}, \quad
	G_n(x,y) := 
		\begin{pmatrix}
			x\idmat{2l+1} \\
			y\idmat{2l+1} \\
			z\idmat{2l+1}
		\end{pmatrix}  \in \C^{3(2l+1)\times(2l+1)}. \label{eqn:clenshawmats2}
\end{align}

@@

\bstodoinline{quad rule on sphere?}

\subsubsection{Function evaluation}\label{subsubsection:clenshaw}

Once again, let $f(x,y,z)$ be a function on the unit sphere $\Omega$ be approximated by its expansion 
\bseqn{
	f(x,y,z) \approx \SHvec(x,y,z)^\top \fvec = \sum^N_{l=0} \SHvecl(x,y,z)^\top \fvec_l,
}
where $\SHvecl (x,y,z), \fvec_l \in \C^{2l+1}$ for each  $l \in \{0,\dots,N\}$, for some coefficients vector $\fvec$ up to degree order $N \in \N$. We can use the Clenshaw algorithm to evaluate this function at a given point $(x,y,z)$ on $\Omega$ \cite{press2007numerical}. The \textit{Clenshaw algorithm} is then as follows:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \bold{0}, \: \vec{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \fvec_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \vec{\xi}_{0}^T \SHvec_0(x,y,z) \equiv \xi_0 Y_0
}
where $\SHvec_0 \equiv \alpha^0_0$.

\subsubsection{Function evaluation}

@@

Since the spherical harmonics are a basis for the surface of the sphere, and the tangent space of the sphere is spanned by the gradient and perpendicular gradient of a scalar function, we have that the gradients and perpendicular gradients of the spherical harmonics 
\bseqnnumber{
	\gradYlm, \quad \gradpYlm := \rvec \times \gradYlm, \label{eqn:VSHs}
}
span the tangent space. Here, $\rvec$ is simply the outward unit normal vector to the surface of the sphere at the point $(x,y,z)$. Importantly, the set $\{\gradYlm, \gradpYlm\}$ form a complete and orthogonal basis for vector valued functions in the tangent space of the sphere and are known as \textit{vector spherical harmonics (VSHs)} \cite{barrera1985vector}. Other definitions for VSHs are used, however it is convenient for deriving explicit sparse relations and operators to use the one described here.

@@

\subsection{Function evaluation}

The Clenshaw algorithm presented in \bsrefsection{subsubsection:clenshaw} can also be used to evaluate a vector-valued function in the tangent space for a point $(x,y,z) \in \Omega$. Let $\uvec$ be a vector-valued function on $\Omega$ whose values are of the form
\bseqn{
	\uvec(x,y,z) = u_\varphi(x,y,z) \: \phivec + u_\theta(x,y,z) \: \thetavec, \quad (x,y,z) \in \Omega
}
for some scalar functions $u^\varphi, u^\theta$ where $\phivec, \thetavec$ are the standard non-radial unit basis vectors on the sphere defined by $\phivec := (\cos \theta \sin \varphi, \sin \theta \sin \varphi, \cos \varphi)^\top$, $\thetavec := (-\sin \theta, \cos \theta, 0)^\top$. Such a function $\uvec$ we refer to as a \textit{vector-valued function in the tangent space of $\Omega$}. We can then approximate this function by its expansion in the VSHs basis, i.e. for some large enough $N \in \N$:
\bseqn{
	\uvec(x,y,z) \approx \VSHvecfull(x,y,z)^\top \uvec^c := \sum^N_{l=0} \sum^l_{m=-l} \big[u_{l,m} \, \gradYlm(x,y,z)  + u^\perp_{l,m} \, \gradpYlm(x,y,z) \big].
}
The clenshaw algorithm to evaluate $\uvec$ at a point $(x,y,z) \in \Omega$ is then as follows
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \bold{0}, \: \vec{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:1 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^\top = (\mathbf{u}^c_n)^\top - \vec{\xi}_{n+1}^\top \Dnt (B_n - G_n) -  \vec{\xi}_{n+2}^\top D^\top_{n+1} C_{n+1} \\
	\quad &\text{3) } \text{Output: } \mathbf{u}(x,y,z) \approx \VSHvecfull_1(x,y,z)^\top \vec{\xi}_{1} \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad = (\vec{\xi}_{1})_1 \gradY_1^{-1} + (\vec{\xi}_{1})_2 \gradpY_1^{-1} + (\vec{\xi}_{1})_3 \gradY_1^{0} \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad + (\vec{\xi}_{1})_4 \gradpY_1^{0} + (\vec{\xi}_{1})_5 \gradY_1^{1} + (\vec{\xi}_{1})_6 \gradpY_1^{1}.
}

@@

\subsection{Example: linear shallow water equations}

@@

\section{Appendix}

\bibliography{../includes/thesis-ben-snowball}

\end{document}

>>>>

\chapter{Spherical harmonics as orthogonal polynomials in three variables}

@@

\section{Defining the spherical harmonics in three variables}

@@

\section{Jacobi matrices}

@@

\begin{proof}
	The recurrence relationship for the Jacobi polynomials satisfies \cite[18.9.1, 18.9.2]{DLMF} 
\bseqn{
	z P^{(m,m)}_{l-m}(z) = \tilde \gamma_{l,m,1} P^{(m,m)}_{l-m-1}(z) + \tilde \gamma_{l,m,2} P^{(m,m)}_{l-m+1}(z),
}
for $k \ge 0, \, m \in \Z$. Hence, the expression for multiplication of the spherical harmonic $\Ylm$ by $z$ is
\bseqn{
	z\,\Ylmxyz &= \clm \eimphi \rho(z)^{|m|} z P^{(|m|,|m|)}_{l-|m|}(z)  \\
		     &= \clm \eimphi \rho(z)^{|m|} \big[ \tilde \gamma_{l,m,1} P^{(|m|,|m|)}_{l-|m|-1}(z) + \tilde \gamma_{l,m,2} P^{(|m|,|m|)}_{l-|m|+1}(z) \big] \\
		     &= \gamma_{l,m,1} Y^{m}_{l-1}(x,y,z) + \gamma_{l,m,2} Y^{m}_{l+1}(x,y,z).
}

For multiplication by $x$ and $y$, we require some further relations for the complex exponential and the Jacobi polynomials. First, recall that
\bseqn{
	\cosphi \, \eimphi &= \frac{1}{2} (e^{i\varphi} + e^{-i\varphi}) \eimphi =  \frac{1}{2} (e^{i(m+1)\varphi} + e^{i(m-1)\varphi}) \\
	\sinphi \, \eimphi &= \frac{1}{2i} (e^{i\varphi} - e^{-i\varphi}) \eimphi =  \frac{-i}{2} (e^{i(m+1)\varphi} - e^{i(m-1)\varphi})
}
Finally, we can express a Jacobi polynomial $P^{(a,b)}_{l-m}(z)$ in terms of an ultraspherical polynomial $C^{(\lambda)}_{l-m}(z)$ \cite[18.7.2]{DLMF}:
\bseqn{
	{l! \: (2m)! \over (l+m)! \: m!} \: C^{(m+1/2)}_{l-m}(z) = P^{(m,m)}_{l-m}(z),
}
where the ultraspherical polynomials satisfy the relations \cite[18.9.7, 18.9.8]{DLMF}:
\bseqn{
	C^{(\lambda)}_n(z) &= {\lambda \over n + \lambda} \: \big[ C^{(\lambda+1)}_n(z) - C^{(\lambda+1)}_{n-2}(z) \big]  \\
	(1-z^2) \, C^{(\lambda)}_n(z) &= {1 \over 4(\lambda-1)(n+\lambda)} \: \big[ (n+2\lambda-2)(n+2\lambda-1) \: C^{(\lambda-1)}_n(z) \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad - (n+1)(n+2) \: C^{(\lambda-1)}_{n+2}(z) \big] ,
}
for $m \in \No$ where $\lambda > -\half$ is some parameter. Thus, combining these, we can write three-term recurrences for the Jacobi polynomials as:
\bseqn{
	P^{(m,m)}_{l-m}(z) &= \tilde \alpha_{l,m,4} P^{(m+1,m+1)}_{l-m}(z) + \tilde \alpha_{l,m,2} P^{(m+1,m+1)}_{l-m-2}(z) \\
	(1-z^2) \, P^{(m,m)}_{l-m}(z) &= \tilde \alpha_{l,m,3} P^{(m-1,m-1)}_{l-m+2}(z) + \tilde \alpha_{l,m,1} P^{(m-1,m-1)}_{l-m}(z),
}
for $l, m \in \No$ s.t. $0 \le m \le l$. Hence, the expressions for multiplication of the spherical harmonic $\Ylm$ by $x$ and $y$ is then:
\bseqn{
	x\,\Ylmxyz &= \clm \cosphi \: \eimphi \: \sintheta \: \rho(z)^{|m|} P^{(|m|,|m|)}_{l-|m|}(z) \\
	&= \half \clm (e^{i(m+1)\varphi} + e^{i(m-1)\varphi}) \rho(z)^{|m|+1} P^{(|m|,|m|)}_{l-|m|}(z) \\
	&= \half \clm e^{i(m+1)\varphi} \rho(z)^{|m|+1} \: \big[ \tilde \alpha_{l,m,4} P^{(|m|+1,|m|+1)}_{l-|m|}(z) + \tilde \alpha_{l,m,2} P^{(|m|+1,|m|+1)}_{l-|m|-2}(z) \big] \\
	&\quad \quad + \half \clm e^{i(m-1)\varphi} \rho(z)^{|m|-1} \: \big[ \tilde \alpha_{l,m,3} P^{(|m|-1,|m|-1)}_{l-|m|+2}(z) + \tilde \alpha_{l,m,1} P^{(|m|-1,|m|-1)}_{l-|m|}(z) \big] \\
	&= \alpha_{l,m,1} Y^{m-1}_{l-1}(x,y,z) +  \alpha_{l,m,2} Y^{m+1}_{l-1}(x,y,z) \\
	&\quad \quad + \alpha_{l,m,3} Y^{m-1}_{l+1}(x,y,z) + \alpha_{l,m,4} Y^{m+1}_{l+1}(x,y,z), \\
	y\,\Ylmxyz &= \clm \sinphi \: \eimphi \: \sintheta \: \rho(z)^{|m|} P^{(|m|,|m|)}_{l-|m|}(z) \\
	&= -\half i \: \clm (e^{i(m+1)\varphi} - e^{i(m-1)\varphi}) \: \rho(z)^{|m|+1} P^{(|m|,|m|)}_{l-|m|}(z) \\
	&= - i \: \big[\Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \big] \\
	&\quad \quad \quad + i \, \big[ \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z) \big], \\
	&= i \: \Big\{ \alpha_{l,m,1} Y^{m-1}_{l-1}(x,y,z) - \alpha_{l,m,2} Y^{m+1}_{l-1}(x,y,z) \\
	&\quad \quad \quad + \alpha_{l,m,3} Y^{m-1}_{l+1}(x,y,z) - \alpha_{l,m,4} Y^{m+1}_{l+1}(x,y,z) \Big\}.
}

@@

\section{Three-term recurrence relation for $\SHvec$}

Three-term recurrence relations for orthogonal polynomials are well established (e.g. \cite[18.9]{DLMF}, \cite{dunkl2014orthogonal}). In a similar vein, we can obtain a recurrence relation for the spherical harmonics by combining each system in (\ref{eqn:jacobimatsdef}).
\bseqn{
	\renewcommand\arraystretch{1.3}
	\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y,z) & A_0 & & \\
		C_1 & B_1-G_1(x,y,z) & \quad A_1 \quad & \\
		& C_2 & B_2 - G_2(x,y,z)  & \ddots \\
		& & \ddots &\ddots
	\end{pmatrix}
	\SHvec(x,y,z)
	=
	\begin{pmatrix}
		Y_0 \\ 0 \\ 0 \\ 0 \\ \vdots
	\end{pmatrix},
}
where we note $Y_0^0(x,y,z) \equiv Y_0 := c_0^0 \: P_0^{(0,0)} \equiv \half {1 \over \sqrt{\pi}}$, and for each $l = 0,1,2\dots$,
\begin{align}
	A_l &:= 
		\begin{pmatrix}
			A_{x,l} \\
			A_{y,l} \\
			A_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l+3)}, \quad
	C_l := 
		\begin{pmatrix}
			C_{x,l} \\
			C_{y,l} \\
			C_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l-1)} \quad (n \ne 0), \label{eqn:clenshawmats1} \\
	B_l &:= 
		\begin{pmatrix}
			B_{x,l} \\
			B_{y,l} \\
			B_{z,l}
		\end{pmatrix} \in \C^{3(2l+1)\times(2l+1)}, \quad
	G_n(x,y) := 
		\begin{pmatrix}
			x\idmat{2l+1} \\
			y\idmat{2l+1} \\
			z\idmat{2l+1}
		\end{pmatrix}  \in \C^{3(2l+1)\times(2l+1)}. \label{eqn:clenshawmats2}
\end{align}

@@

\section{Computational aspects}

Once again, let $f(x,y,z)$ be a function on the unit sphere $\Omega$ be approximated by its expansion 
\bseqn{
	f(x,y,z) \approx \SHvec(x,y,z)^\top \fvec = \sum^N_{l=0} \SHvecl(x,y,z)^\top \fvec_l =  \sum^N_{l=0}  \sum^l_{m=-l} f_{l,m} \: \Ylmxyz,
}
where $\SHvecl (x,y,z), \fvec_l \in \C^{2l+1}$ for each $l \in \{0,\dots,N\}$, for some coefficients vector $\fvec = (f_{l,m})$ up to degree order $N \in \N$.

\subsection{Obtaining coefficients}

\bstodoinline{obtaining coeffs/ transform}
In spectral space, we wish to work only with vectors of coefficients for the expansion of a function, to which we can apply operator matrices to that represent differential or other operations. Naturally, we of course need a way to obtain the coefficients $f_{l,m}$. The coefficients can be calculated via the integral
\bseqn{
	f_{l,m} = \int_\Omega f(\costheta \sinphi, \sintheta \sinphi, \cosphi) \: \Ylmfull^* \: \sinphi \: \D \varphi \: \D \theta
}
using the orthonormality of the spherical harmonics. Methods to calculate these coefficients exist, known as spectral transforms, and are well established for the spherical harmonics (see e.g. \cite{mohlenkamp1999fast, suda2002fast}).

\subsection{Function evaluation}\label{subsubsection:clenshaw}

We can use the Clenshaw algorithm to evaluate this function at a given point $(x,y,z)$ on $\Omega$ \cite{press2007numerical}. The \textit{Clenshaw algorithm} is then as follows:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \bold{0}, \: \vec{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \fvec_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \vec{\xi}_{0}^T \SHvec_0(x,y,z) \equiv \xi_0 Y_0.
}

\subsection{Variable coefficients as operators}

@@

Since the spherical harmonics are a basis for the surface of the sphere, and the tangent space of the sphere is spanned by the gradient and perpendicular gradient of a scalar function, we have that the gradients and perpendicular gradients of the spherical harmonics 
\bseqnnumber{
	\gradYlm, \quad \gradpYlm := \rvec \times \gradYlm, \label{eqn:VSHs}
}
span the tangent space. Here, $\rvec$ is simply the outward unit normal vector to the surface of the sphere at the point $(x,y,z)$. Importantly, the set $\{\gradYlm, \gradpYlm\}$ form a complete and orthogonal basis for vector valued functions in the tangent space of the sphere and are known as \textit{vector spherical harmonics (VSHs)} \cite{barrera1985vector}. Vector spherical harmonics have been widely used in electrostatics (e.g. \cite{barrera1985vector}), electrodynamics (e.g. \cite{carrascal1991vector}), and of course fluid dynamics including weather and climate modelling (e.g. \cite{moses1974use, fengler2005nonlinear, swarztrauber1996spectral}). Other definitions for VSHs are used (e.g. \cite{hill1954theory}), however it is convenient for deriving explicit sparse relations and operators to use the one described here.

@@

Let $\uvec$ be a vector-valued function on $\Omega$ whose values are of the form
\bseqn{
	\uvec(x,y,z) = u_\varphi(x,y,z) \: \phivec + u_\theta(x,y,z) \: \thetavec, \quad (x,y,z) \in \Omega
}
for some scalar functions $u^\varphi, u^\theta$ where $\phivec, \thetavec$ are the standard non-radial unit basis vectors on the sphere defined by $\phivec := (\cos \theta \sin \varphi, \sin \theta \sin \varphi, \cos \varphi)^\top$, $\thetavec := (-\sin \theta, \cos \theta, 0)^\top$. Such a function $\uvec$ we refer to as a \textit{vector-valued function in the tangent space of $\Omega$}. We can then approximate this function by its expansion in the VSH basis, i.e. for some large enough $N \in \N$:
\bseqn{
	\uvec(x,y,z) \approx \VSHvecfull(x,y,z)^\top \uvec^c := \sum^N_{l=0} \sum^l_{m=-l} \big[u_{l,m} \, \gradYlm(x,y,z)  + u^\perp_{l,m} \, \gradpYlm(x,y,z) \big].
}

\subsubsection{Obtaining coefficients}
The development of vector spherical harmonic transforms is well established \cite{swarztrauber1993vector}, and continues today (see \cite{gia2019favest}). \bstodoinline{VSH Transform}

\subsection{Function evaluation}
The Clenshaw algorithm presented in \bsrefsection{subsubsection:clenshaw} can also be used to evaluate a vector-valued function in the tangent space for a point $(x,y,z) \in \Omega$:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \bold{0}, \: \vec{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:1 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^\top = (\mathbf{u}^c_n)^\top - \vec{\xi}_{n+1}^\top \Dnt (B_n - G_n) -  \vec{\xi}_{n+2}^\top D^\top_{n+1} C_{n+1} \\
	\quad &\text{3) } \text{Output: } \mathbf{u}(x,y,z) \approx \VSHvecfull_1(x,y,z)^\top \vec{\xi}_{1} \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad = (\vec{\xi}_{1})_1 \gradY_1^{-1} + (\vec{\xi}_{1})_2 \gradpY_1^{-1} + (\vec{\xi}_{1})_3 \gradY_1^{0} \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad + (\vec{\xi}_{1})_4 \gradpY_1^{0} + (\vec{\xi}_{1})_5 \gradY_1^{1} + (\vec{\xi}_{1})_6 \gradpY_1^{1}.
}

@@

\section{Example: linear shallow water equations}

@@
